Outline flowCore paper

### 2008-02-20 paper/outline.tex
updated paper draft
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@30063 bc3139a8-67e5-0310-9ffc-ced21a209358

\begin{document}
\maketitle

@@

\bigskip
FH put the following very rough structure up for discussion:

@@

%%\subsection{High throughput flow}
A brief introduction to the flow technology and high throughput flow
in particular.
{\bf RB:} Given the proposed journal, an introduction to flow technology isn't
required.

\textit{introduction or abstract (first draft)}

Traditionally, flow cytometry has been a tube-based technique limited
to small-scale laboratory and clinical studies.  High throughput
methods for flow cytometry have recently been developed and are now
used in both basic and clinical research. Today, the need for flexible
and well structured computatioonal tools to efficiently handle and
analyze FC-HCS has considerably increased. Indeed, the amount of
information generated by these technology must be stored, managed, and
needs to be summarized in order to be accessible to the
researcher. However, the absence of any shared research platform which
bioinformaticians, computer scientists, and statisticians can use to
develop standard or novel methods for flow cytometric has become
problematic. We believe that the Open Source statistical software R in
conjuction with the Bioconductor Project can fill this gap.  In this
paper, we propose R data structures to handle flow cytometry data
through the main steps of importing, storing, assessing and
preprocessing data from FCM experiments. These methods, along with a
number of specialized functions for compensation, transformation and
filtering and other preprocessing steps, are implemented in the newly
developped Bioconductor package flowCore.  We envision flowCore to be
a solid foundation for researchers to efficiently handle FC-HCS data
and for future development of new methods and tools for the analysis
of flow cytometry.

\textit{end --- introduction (first draft)}

\subsection{R and Bioconductor}

\textit{ example text}

R is a solid statistical framework/environment, particularly geared
towards the analysis the large numeric datasets like those arising
from high throughput experiments such as FC-HCS. Indeed, this open
source software already offers a wide range of statistical and
visualization methods developed for various field of application, many
of them relevant also to FC-HCS data analysis. 
%% FIXME: either cite some of our paper or drop this sentence. 
%% I think it it too redundant with the one above it.
In addition, R is a research platform which
bioinformaticians, computer scientists, and statisticians are using to
develop new tools for statistical data analyis. For biological
application, the Bioconductor project has become the standard toolset
\citep{gentleman2006bos}...

NLM: reference to microarray ?
\textit{end--- example text}

\subsection{Existing data standards and conventions}

\textit{ example text}

How does the flow data world look like FCS Other flow software (what
other software are handling multiple FCS files all together
). Somebody more familiar with flowJo and friends should add to
this...

FCS -> we can read these files

\textit{end --- example text}

@@

{\bf PH:} I think the Frame/Set concept is pretty interesting and
powerful. Was it originated for this project? The ability to program
sets flowCore aside from all other flow software applications, I
think. I think there is something really important and valuable about
how the data structures are ogranized. I'm not sure how to write that
in a way to appeal to Cytometry A readers, but I think we could work
this out. Second, I think we have to consider what would be the
motiviation for 
\begin{itemize}
\item why the journal would publish the paper 
\item why someone would read the paper, and
\item why someone would want to try the software
\end{itemize}

\section{Basic data structures}

\textit{example Text}

\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of flow cytometry (or similar) data. This is accomplished
through a data model very similar to that adopted by other
Bioconductor packages used for gene expression analysis and thus
familiar to most Bioconductor users \cite{}.  The data structure is
designed to combine several differents sources of information into a
single convenient structure: (\textit{e.g.,} measurements from a
combination of samples, metadata describing samples in the
experiments).

What are the parts we need to keep together

flowCore does the maintenance, the user doesn't have to care about
that

The concept of sets

Figure: A diagram of the data structures

\textit{end Text}

{\bf BE:} For this section I think we'd want to describe the philosophy that
underlies the data structure rather than a technical description of
the data structure since they weren't chosen arbitrarily. Like any
other flow tool we start with a collection of events, a flowFrame.
However, we would like to do tasks consistently across several frames
so we introduce a flowSet (a set because there is generally no
inherent ordering) that manages the consistent application of
operations and so on.

\subsection{demands on flow data structures}

numeric raw data matrices
instrument-specific meta data
experiment meta data
memory issues, scale of data (tubes, plates)

\subsection{flowFrame}

\textit{example Text}

The basic unit of manipulation in \Rpackage{flowCore} is the
\Rclass{flowFrame}, which corresponds roughly with a single ``FCS''
file exported from the flow cytometer's acquisition software. The
\Robject{flowFrame} object store the event-level information along
with a number of additional metadata. The event information, stored as a
single matrix, is easily accessed and manipulated (see
\textit{Supplementary Materials}). All available information regarding
detectors and stains is drawn from the FCS files and attached to the frame.

\textit{end Text}

{\bf NLM:} Should be more abstract ?

\subsection{flowSet} 
experiments, multiple FCS files, plates

\section{Standard flow operations}
{\bf BE:} We would also like to abstractly describe these operations. This is
somewhat more difficult, especially since we'd like other packages to
be able to extend these operations and introduce their own as
first-class citizens in the analysis. Additionally, we would like to
factor out as much bookkeeping as possible to allow authors to focus
on the actual operation rather than interfacing with flowSets versus
flowFrames and so on. We can then describe the general classes of
operations and how they interact.

\subsection{transformations}
Different flavors, R's own transformation infrastructure allows for
arbitrary transforms, we have implemented the commonly used one.

\subsection{gating}
gate definitions
capturing gating schemes, workflow
gate summary statistics

\section{Application, Outlook}
Challenges in high throughput cytometry
(QA, automated gating, class discovery)

flowCore as toolbox

already in devel 'plug-in' packages (Visualization, QA, stats, ...)

>>>>

\begin{document}

@@

\bigskip

\maketitle

@@

Traditionally, flow cytometry (FC) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry (FC-HCS) have recently been
developed and are now used in both basic and clinical research. Today,
the need for flexible and well structured computational tools to
efficiently handle and analyze FC-HCS has increased
considerably. Indeed, the amount of information generated by these
technologies must be stored, managed, and needs to be summarized in
order to be accessible to the researcher. However, the absence of any
shared research platform which bioinformaticians, computer scientists,
and statisticians can utilze to develop novel or standard methods for
flow cytometric has become problematic. We believe that the open
source statistical software R in conjuction with the Bioconductor
Project can fill this gap.  In this paper, we propose R data
structures to handle flow cytometry data through the main steps of
importing, storing, assessing and preprocessing data from FC-HCS
experiments. These structures, along with a range of specialized
methods for compensation, transformation, gating and additional
preprocessing steps, are implemented in the newly developed
Bioconductor package flowCore.  We envision flowCore to be a solid
foundation for researchers to efficiently handle FC-HCS data and for
future development of new methods and tools for their coherent
analysis.

\subsection{R and Bioconductor}
R is a solid statistical programing environment, particularly geared
towards the analysis of large numeric datasets typical to arise from
high throughput experiments such as microarray gene expression anlysis
or FC-HCS. Today, the R open source project offers a wide range of
statistical and visualization methods developed for various fields of
application. In addition, R is a research platform which
bioinformaticians, computer scientists, and statisticians routinely
use to develop novel tools and methodology for statistical data
analyis. For biological applications, the Bioconductor project has
become the standard toolset \citep{gentleman2006bos} to process data
spanning a diversity of research fields from genomics to proteomics to
complex cell biology. This shared platform proved to be especially
beneficial for the development of analyis routines for microaray gene
expression data, and by now many of these innovations have been
incoorporated into comercial software products, thus greatly improving
the quality of gene expression data analysis. Shared data structures
have been shown to be essential for this kind of collaborative effort.

\subsection{Existing data standards and conventions}
Currently, data from flow cytometry experiments are stored in single
files following a well-defined data standard (FCS). This standard
tries to capture as much of the accompanying meta information as
possible, most of which is produced by the measurement instrument or
has to be supplied manually by the operator upon experiment
setup. However, the recent developments in high-throughput flow
cytometry are shifting the focus of interest away from single-tube
based measurements towards large and complex experimental designs with
dozens of co-variates and influencing factors. Self-contained data
structures are a prerequesite to allow for coordinated analysis of
such data sets. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programatical access. In addition, the
closed-source nature of most of these products makes it virtually
impossible to connect and integrated these data.

Traditionally, the majority of flow cytometry experiments have been
analysed manually by direct data inspection or very basic comparisons
of summary statistics. We believe that these approaches do not give
justice to the highly complex nature of FC data, essentially
discarding many of the fundamental aspects such as data distribution
or high-dimensionality. In addition, the subjective character of
manual analyses are a major obstacle for reproducibility. For FC-HCS
data, unassisted manual inspection is no longer feasible, and robust
statistical methods need to be developed to point the investigator to
the intersting aspects of the data or to potential problems. While the
expert knowledge of immunologists and researchers remains crucial for
the understanding of FC data, we believe that tight colaboration
with other reserach fields such as statistics and computer science can
greatly improve the relevance of flow cytomtry in today's
high-throughput biological sciences.

\textit{How does the flow data world look like? Other flow software
  (what other software are handling multiple FCS files at
  all). Somebody more familiar with flowJo and friends should add to
  this... Also somebody should fill in a couple of sentences about the
  gating-ML etc. stuff here}

The flowCore framework presented here offers import functionality for
both the raw data FCS files along with the complete set of
file-specific meta information as well as all the information that is
stored in one of the gating-, compensation- or transformation-ML
standards. Thus, it is possible to intergrate flowCore in existing
work flows and to communicate with virtually every other FC tool that
adheres to these standards.

@@

\section{Basic data structures}
\subsection{flowFrame}
flowCore's primary task is the representation and basic manipulation
of flow cytometry (or similar) data. This is accomplished through a
data model very close to that adopted by other Bioconductor packages
used for gene expression analysis and thus familiar to most
Bioconductor users. Similar to any other flow tool, the basic unit of
organisation in flowCore is a collection of events. We call the
structure that hold this data a flowFrame. All relevant information
for the collection are stored along with the matrix of raw data values
and can be accessed programatically. Most prominently, these are
descriptors of the stains used in the experiment and the respective
measurement channels, information about compensation performed at the
instrument side and any additional keywords the user deems to be
important to annotate the data. A number of quality checks are
performed during the creation of a flowFrame to ensure data
integrety. There is an extended interface that allows for the
standardized interaction with the flow data, thus enabling the
development of more specialized tools for specific tasks of the data
analysis process.

\subsection{flowSet}
In high-throughput flow cytometry, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. flowSets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flowFrames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of mixups and errors that are
all too common when dealing with complex data in the ususal way of
more or less ad hoc spreadsheats that lack a controlled structure. In
general, there is no native ordering to a flowSet, which in a way
reflects the still predominant application of tube-based experiments
rather than experiments following a plate-design. However, the flowSet
structure can be readily extended to also accomodate such spacial
information \textit{(reference)}

\textit{Figure: A diagram of the data structures and the basic
  operations (Figure1.ppt)}

{\bf PH:} I think the Frame/Set concept is pretty interesting and
powerful. Was it originated for this project? The ability to program
sets flowCore aside from all other flow software applications, I
think. I think there is something really important and valuable about
how the data structures are ogranized. I'm not sure how to write that
in a way to appeal to Cytometry A readers, but I think we could work
this out.

\section{Standard flow operations}
The basic operations in flow cytometry analysis are almost always the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flowFrames as well as on entire
flowSets. While transformation and to a certain extent also
compensation are fairly routine operations with only limited potential
for improvement, being able to implement new methodology for gating of
flow cytometry data and extending the capabilies of flowCore through
the usual ways of object oriented programming is a feature, that
clearly sets our framework apart from all the other flow toolsets
around. By factoring out as much of the bookkeeping as possible,
authors can focus on the actual operations rather then having to deal
with the teadious details of data integration and access. Third party
methods implemented by others can act on their own as first-class
citizens in the analysis, without breaking the workflow or the basic
infrastructure. This design allows for the straight forward extension
of flowCore's capabilities and already fostered the development of a
number of valuable improvements. \textit{(cite stuff here)}

\subsection{transformations}
All the major transformations that are routinely used in flow
cytometry analysis have been implemented in flowCore. Furthermore, the
design of the R language makes it easy to define arbitrary functions
to apply to the data of individual flowFrames or entrire flowSets,
respectively. Proper transformation have been proven beneficial for
both data visualization and modelling \textit{cite flowClust paper}.

\subsection{gating}
In flowCore, gates are represented as individual objects that can be
extended in an object oriented manner. All basic gate types such as
rectangular gates, ellipses and polygons are implemented as part of
the base framework. In addition, we introduce the notion of
data-driven gates, for which the necessary parameters are computed
based on the properties of the underlying data, for instance
distribution or density estimates. The ability to programatically
access gates is a prerequesite for automated or semi-automated
gating. By utilizing a unified interface for all different types of
gates, the user is able to subset data sets as well as create summary
statistics, for instance of proportions of events falling in single
gates or combinations of multiple gates. The definition of gates in
flowCore follows the proposed gating-ML standard, thus any flowCore
gating strategy can be reproduced by any other software that also
adheres to the standard.

\section{Related flow packages}
In addition to the flowCore package that offers basic infrastructure
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of flow cytomtery data
analysis. The flowViz package provides sophisticated data
visualization tools that make use of multivariate trellis
plotting. These functions can be used to quickly generate customized
plots for extended cytometry data sets for both direct data inspection
and quality control. The flowQ package offers more advanced quality
assurance (QA) methodology and a framework to create interactive
reports of QA results. Most of the flowUtils content deals with data
import and output from and to flowCore and the flow-cytometry specific
standard markup languages. Finally, flowStats provides statistical
methods that are relevant in the context of flow cytometry data
analysis.

\textit{A diagram of the packages and what they offer}

\section{Application, Outlook}
\textit{This is the most important part, we need to show nice examples
  of use cases, limit ourselfs to the basics and show what we can do
  that nobody else can do at the moment. Need at lot of input here
  from everybody. Also for the things that go in here we should
  provide further details in the Supplementary Materials.}
